---
title: "Visualize"
---

## Module Learning Objectives

By the end of this module, you will be able to:

- <u>Create</u> a baseline plot with `ggplot2`'s `ggplot` function along with the `aes` helper function.
- <u>Add</u> desired geometries to your baseline plot with the `geom_...` family of functions
- <u>Explain</u> the advantage of using the `+` operator to add together multiple plot elements
- <u>Use</u> `ggplot2`'s `labs` and `scale_fill_manual` helper functions to customize your plot's labels and colors 
- <u>Differentiate</u> `aes`' `fill` and `color` arguments 
- <u>Generate</u> separate plots based on grouping variable(s) with `ggplot2`'s `facet_grid` function
- <u>Design</u> a custom format for your plots using `ggplot2`'s `theme` function

```{r libraries-visualize, echo = F, message = F}
library(tidyverse); library(palmerpenguins)
```

## `ggplot2` Overview

While the bulk of the `tidyverse` is focused on modifying a given data object, `ggplot2` is also a package in the `tidyverse` that is more concerned with--intuitively enough--*plotting* tidy data. `ggplot2` does share some syntax with the functions and packages that we've discussed so far but it also introduces some new elements that we'll discuss as we encounter them.

## Creating a Plot

To create the foundation for your plot, we'll use the `ggplot` function (note that the package is `ggplot2` while the function name lacks the "2"). This function allows you to globally define the data object you're using as well as which variable(s) should be mapped to x and y axes as well as aesthetic parameters--for example, which groups outline or fill color should be inherited from, etc. The `ggplot` function relies on an aesthetics helper function named `aes`. We'll give `ggplot` the data name and `aes`, while we'll pass all relevant variables to `aes` directly. This may be easier to follow once we've covered an example so let's do that now!

:::callout-note
## Example

Just like the preceding chapters, let's use the `penguins` dataset to demonstrate this plot. Let's create the foundation of a graph that has `year` on the y-axis and `body_mass_g` in the x-axis.

The `ggplot` function wants both a `data` and `mapping` argument that we'll specify here for clarity but will exclude going forward. `mapping` expects the `aes` function that in turn defines all your variable placements.

```{r ggplot-no-geom, message = F}
ggplot(data = penguins, mapping = aes(x = year, y = body_mass_g))
```

It may not look like it, but this was successful! The `ggplot` function creates the first "layer" of the plot including raw axis titles and tick marks but when creating a plot with the `ggplot2` package, choosing *which* plot type is actually done with a separate class of functions called **"geometries".**
:::

## Choosing a Plot Type

Now that we have a baseline plot, we can add desired geometries using the `geom_...` family of functions. Broadly speaking, there is one `geom_...` for every possible way of plotting your data. Want to make a scatter plot? Use `geom_point`. Bar plot? `geom_bar`. Add a best-fit line? `geom_smooth`. When you first begin making plots with `ggplot2` you will likely have to Google which `geom_...` you want (that was certainly what the creators of this workshop did when we started out!) but over time you'll remember them more and more clearly.

### Geometry Aside No. 1 - Adding Plot Elements

You may have noticed that the core plot is built with `ggplot` and `aes` but each subsequent component is added with one of the `geom_...` functions and realized the gap we haven't talked about yet: how do we combine these separate lines of code? The answer is part of what makes `ggplot` different from the rest of the `tidyverse`. In the rest of the `tidyverse` we chain together multiple lines of code with the `%>%` operator, however, **in `ggplot2` we use `+` to combine separate lines of code.**

This has a distinct advantage that we'll discuss later but we'll use the `+` in the following example to show its use.

### `geom_...` Example: Adding a Geometry

:::callout-note
## Example

Let's re-create our `year` by `body_mass_g` plot but let's make it a scatter plot by adding `geom_point`.

```{r geom-single, warning = F, message = F}
ggplot(data = penguins, mapping = aes(x = year, y = body_mass_g)) +
  geom_point()
```

Because `aes` gets all of the variable mapping information, we don't need to give *anything* to `geom_point`! This makes adding multiple geometries much easier than if we had to re-specify the variables separately in every geometry!
:::

### `geom_...` Example: Adding Multiple Geometries

:::callout-note
## Example

Let's add a best-fit line to this graph to demonstrate how multiple geometries are added.

```{r geom-multiple, warning = F, message = F}
ggplot(data = penguins, mapping = aes(x = year, y = body_mass_g)) +
  geom_point() +
  geom_smooth()
```

Not a terribly informative graph but the code aptly demonstrates how several geometries can be layered.
:::

### Geomtery Aside No. 2 - Order Matters

The heading says it all: *order matters!* The order that you add `geom_...`s to your plot (using `+`) determines which geometries are "above" or "in front of" others. This is a desirable behavior for plots with multiple geometries but is something to keep in mind! Let's cover an example to clarify this.

### `geom_...` Example: Order

:::callout-note
## Example

Let's say we want to make a `ggplot` of bill depth within each penguin species and we want both a boxplot and the points that make up the boxplot. This involves using both the `geom_point` and `geom_boxplot` geometries but order will be crucial! To demonstrate our point better, we'll also add a new argument to our `aes`: `fill`. `fill` defines which variable should be used to fill in open spaces (like bar or box plots). Let's create our plot!

```{r geom-order-1, warning = F, message = F}
ggplot(penguins, aes(x = species, y = bill_depth_mm, fill = species)) +
  geom_point() +
  geom_boxplot()
```

Our plot has all the right elements, but the box plots are covering up the points between the ends of the box (the 25th and 75th quartiles). If we change the geometry order we'll get our points "in front of" the box plots.

```{r geom-order-2, warning = F, message = F}
ggplot(penguins, aes(x = species, y = bill_depth_mm, fill = species)) +
  geom_boxplot() +
  geom_point()
```

Now our points can be seen even when they overlap with the box plot. Note that the order of geometries doesn't affect any overlap in either the x or y axis. In our above plot you'll see that our points are so densely stacked that it is difficult to see the whiskers' extent. We must change geometries in order to fix overlap issues like this. Fortunately, we can use `geom_jitter` instead of `geom_point` to make a scatter plot where all points are jittered in the x-axis to make them easier to see.

```{r geom-order-3, warning = F, message = F}
ggplot(penguins, aes(x = species, y = bill_depth_mm, fill = species)) +
  geom_boxplot() +
  # The width argument lets us specify how far apart (in the x-axis) we want points to jitter
  geom_jitter(width = 0.25)
```

Great! Now we can see both the points and the box plots because we've achieved **(1) the order of geometries** and **(2) the type of geometries** that we need in this case.
:::

### Challenge: Geometries

:::callout-important
## Your Turn!

Using `ggplot2`, create a scatter plot of `bill_depth_mm` against `body_mass_g`.
:::

## Advantage of the `+`

When we first introduced the `+` for adding together multiple plot elements we said that it had a key advantage but were vague about what exactly that benefit was. **The `+` allows you to build plots stepwise by adding elements to an object**. This can be extremely useful when you're experimenting with a plot as you could make several different objects, each with increasingly more plot elements, and allow all of them to inherit the same fundamental architecture.

For users who are writing their own functions, **the `+` allows you to build plots using conditions specified by arguments**. Say you want to write a function that always makes the same type of `ggplot` but you want to allow users to choose whether a best-fit line is added. Because of the `+` you can do exactly this!

Let's cover a quick example of the value of stepwise plot creation.

### Stepwise Plot Addition Example

:::callout-note
## Example

Let's create a plot of bill length against bill depth and color the points by penguin species. First, we create the foundation of our plot without any geometries and assign it to an object using the `<-` operator.

```{r stepwise-1, warning = F, message = F}
plot_v1 <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species))
plot_v1
```

Now let's add points to this blank plot using `geom_point` (in this case we *do not* want `geom_jitter` because it changes the x-position of points).

```{r stepwise-2, warning = F, message = F}
plot_v2 <- plot_v1 + 
  geom_point()
plot_v2
```

Now we can further add a linear regression line (`geom_smooth(method = "lm")`) for each species' bill depth vs. length relationship. Because we colored by species earlier, `geom_smooth` will automatically create separate regression lines! Pretty cool, right?

```{r stepwise-3, warning = F, message = F}
plot_v3 <- plot_v2 + 
  geom_smooth(method = "lm")
plot_v3
```

Great! We could have added the `ggplot` and `geom_point` and `geom_smooth` lines together at the same time but the `+` allows us to build graphics step-by-step if we so desire.
:::

## Customizing Plots

There are too many helper functions included in `ggplot2` to cover them all here but we can cover two that you may find particularly helpful: `labs` and `scale_fill_manual`.

### Customizing - Labels

`labs` is `ggplot2`'s one-stop shop solution to customizing your axis labels and plot title. Add `labs` in the same way you would add a geometry to add custom axis labels to your plot. This is especially useful if your column names are highly abbreviated or otherwise difficult for those outside of your group to interpret.

### `labs` Example: Adding Labels

:::callout-note
## Example

Let's make a plot of the number of penguins per island per species with more informative axis labels and a plot title. We'll need to calculate that quickly but we can leverage the tools we practiced in the "Summarize" chapter to do that.

```{r labs-prep, warning = F, message = F}
# Count penguins
penguin_count <- penguins %>%
  dplyr::group_by(species, island) %>%
  dplyr::summarize(total = dplyr::n()) %>%
  dplyr::ungroup()

# Take a look at that object
dplyr::glimpse(penguin_count)
```

Now that we have that object, let's make a bar plot of penguins on each island and let's color by penguin species.

```{r labs-actual, warning = F, message = F}
ggplot(penguin_count, aes(x = island, y = total, fill = species)) +
  # `stat` argument defines whether you've already done the calculation or you want `ggplot` to try to do it itself
  geom_bar(stat = 'identity') +
  labs(x = "Island Name",
       y = "Penguins Measured",
       title = "Number Penguins Counted per Island and Species")
```

We now have much more human-readable axis labels and an informative title! `labs` can also be used to add a subtitle, caption, or even alt text (alternative text is special text embedded in an image that is useful for people who require a screen reader to describe visual elements). See `?labs` for more information on the other labels it controls.
:::

### Customizing - Using Custom Colors

As we've progressed through this training you may have noticed the classic `ggplot2` default colors (the first three are a mild red, blue, and green). It is often the case however that we want to choose our own colors; this is where `scale_fill_manual` becomes needed.

This function is added in the same way that the `geom_...` and `labs` functions are added and requires a vector of the colors that you want to use instead of the defaults. There are two key conditions to keep in mind when specifying your colors. First, **you must supply as many colors as there are groups to color _by_**; `ggplot2` will not fill in defaults if you supply too few colors and won't guess which to drop if you give too many. Second, **the order that you provide colors in must be correct** OR **you must "name" each color to assign it to a specific group**. To "name" the color, use the following syntax: `c(groupA = color1, groupB = color2, ...`. Order can be a simple vector but it is often nice to be explicit about which group should be which color.

Let's return to our bar plot above and change the colors used for our bar plot.

### `scale_fill_manual` Example: Specify Fill

:::callout-note
## Example

Let's re-create the plot from the previous example and specify the colors manually using `scale_fill_manual`

```{r scale-fill, warnings = F, message = F}
# Create the plot
ggplot(penguin_count, aes(x = island, y = total, fill = species)) +
  geom_bar(stat = 'identity') +
  # Specify labels
  labs(x = "Island Name", y = "Penguins Measured") +
  # Choose colors!
  scale_fill_manual(values = c("red", "orange", "yellow"))
```

Two quick notes before we move on:

1. You can use "hexadecimal codes" for colors (sort of a machine-readable way of specifying colors) in this function. We find the website [colorbrewer2.org](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) particularly useful because it groups colors into palettes and includes a "colorblind safe" check box that limits your options to only those that will be accessible to all viewers. You can also download the `RColorBrewer` package to easily incorporate its many color palettes into your visualizations!

2. Feel free to check out [NCEAS' R color cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf) for more color options.
:::

### Customizing - `color` versus `fill`

One nuance of `ggplot2` aesthetics that we should cover before continuing is the difference between "color" and "fill". If used in the `aes` function, both of these at first glance seem to dictate what we might think of as "plot color." In the example above, we used `scale_fill_manual` to specify the color of the bars in our bar plot.

Put clearly: `ggplot2` defines "fill" versus "color" as follows:

- `fill` = controls **color inside** of a hollow shape
- `color` = controls **color on the _outside_** of a hollow shape *or* the **entire color of a solid shape**
    - From `ggplot2`'s point of view, a solid shape is essentially a shape that is all outline and both real outlines and non-hollow shapes fall under `color`'s jurisdiction

Let's make some example plots to demonstrate this distinction.

:::callout-note
## Example

Let's make a boxplot of `bill_length_mm` across penguin `species` and set `fill` to `species`.

```{r color-vs-fill-1, message = F, warning = F}
ggplot(penguins, aes(x = species, y = bill_length_mm,
                     fill = species)) +
  geom_boxplot()
```

Notice how the *interior* of the boxes is colored by species? Watch what happens when we make the same plot again but change `fill = ` to `color = ` in the `aes` function:

```{r color-vs-fill-2, message = F, warning = F}
ggplot(penguins, aes(x = species, y = bill_length_mm,
                     color = species)) +
  geom_boxplot()
```

See how the interior of the box plots now defaults to white but the outline edges become colored by `species`? Let's demonstrate this again with a solid object to show `fill` versus `color` in another context.

Let's create a scatterplot of `bill_length_mm` by `bill_depth_mm` and `color` by `species`

```{r color-vs-fill-3, message = F, warning = F}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm,
                     color = species)) +
  geom_point()
```

The default point shape for a `ggplot2` scatterplot is a solid dot so the points are now fully colored by `species` (see `?pch` for other point options). What if we changed `color = ` to `fill = ` though?

```{r color-vs-fill-4, message = F, warning = F}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm,
                     fill = species)) +
  geom_point()
```

Interestingly, `ggplot2` knows that you're trying to differentiate by `species` so it still returns a legend but unfortunately all three species' points defaulted to black so we can't see the difference! If we change the points to be hollow however...

```{r color-vs-fill-5, message = F, warning = F}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm,
                     fill = species)) +
  geom_point(shape = 21)
```

We now get our points *filled* with the correct color with a black outline to each point!

If you ever try to manually specify one or the other and your plot looks the same, chances are you're mis-specifying `color` for `fill` or vice versa. Double check that and you will often resolve the issue
:::

### Challenge: Customizing

:::callout-important
## Your Turn!

Using `ggplot2`, create a scatter plot of `bill_depth_mm` against `body_mass_g` **and** where the points are colored based on the `sex` of the penguin. Make female penguins' points **<span style="color:red">red</span>** and male penguins' points **<span style="color:blue">blue</span>**. Also, give both axes manually-specified titles (i.e., not the raw column names!).
:::

## Separating Plots by a Variable

Sometimes it is useful to create a plot but separate out the data by one of the columns. `ggplot2` includes the `facet_...` family of functions to accomplish this. `facet`ing a plot creates several panels that have the same labels and data but separated by whatever variable(s) you give to the `facet_...` function in question.

### `facet_grid` Example

:::callout-note
## Example

Let's return to our multi-colored bill length vs. bill depth plot to demonstrate `facet`ing in action. We'll create the same plot but this time we'll `facet` by island so that the penguins of each island are included in separate panes.

The syntax of `facet_grid` is as follows: `[column for columns of plots] ~ [column for rows of plots]`. If you only want rows *or* columns of your plot panes, replace that side of the `~` with a period.

```{r facet, warning = F, message = F}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Let's also add better axis labels
  labs(x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  # Now facet by island
  facet_grid(. ~ island)
```

Note that the values within `island` that make up each pane are included in the dark gray bar at the top of each pane. This behavior is automatic and makes your separated plots much more interpretable.
:::

### Challenge: `facet_grid`

:::callout-important
## Your Turn!

Using the plot you created for the previous challenge (`bill_depth_mm` vs. `body_mass_g` and `color = sex`, etc.), facet by `species` of penguin. Note that you can choose whether you want your plot panels stacked vertically or horizontally.
:::

## Changing Thematic Properties

In all of our preceding examples the plots have a characteristic `ggplot2` "feel" where they all have gray backgrounds with white grid lines of varying thickness depending on whether they are minor or major. If `color` or `fill` is specified in `aes` the legend is on the right and in all plots the axis labels and tick marks are (in our opinion) a very small font size.

`ggplot2` placed all of the power to modify any of these parameters into a single, comprehensive function: `theme`. `theme` allows users to specify nearly any formatting component and modify it as they desire. Just like the `geom_...` functions, Google will very much be your friend as you try to remind yourself about the breadth of possibility within `theme` but over time you will become more comfortable and confident!

Let's cover an example to demonstrate a few components of `theme`.

### `theme` Example: Manual Specification

:::callout-note
## Example

Let's create a plot of bill length versus flipper length and colored by penguin species.

```{r theme-1, warning = F, message = F}
core_plot <- ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm,
                                  color = species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Let's also add better axis labels
  labs(x = "Flipper Length (mm)", y = "Bill Length (mm)")

core_plot
```

Now that we have this plot, let's change three things using theme:
1. Increase axis text size
2. Increase axis *title* text size
3. Move the legend into the bottom right of the plot

To accomplish this, we'll add a `theme` to the `core_plot` object that we created above and specify each of those changes within that function.

```{r theme-2, warning = F, message = F}
core_plot +
  # Add `theme` to the plot
  theme(
    # Tell it the X/Y coordinates of the legend
    legend.position = c(0.9, 0.15),
    # Now change the text size of the axis ticks and labels
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14) )
```

Note that for certain parameters, you must also specify one of the `element_...` family of functions. `element_text` governs all things related to text (i.e., size, font, etc.) while `element_rect` changes things related to background boxes of plot components. There are other `element_...` functions that you'll encounter as you use Google and Stack Overflow (a coding help website Google will frequently identify when you search for error explanations) so rest assured that you'll be able to find support documentation for these when you need it.
:::

### `theme` Example: Pre-Packaged Themes

If that all seemed pretty involved, don't worry! `ggplot2` comes with several `theme_...` functions that are pre-built and modify a lot of these parameters in desirable ways without necessitating you doing a deep dive into Google. For instance, `theme_bw` or `theme_classic` are both great options that we'll demonstrate below.

:::callout-note
## Example

Let's re-use the `core_plot` object we built above to quickly demonstrate `theme_bw` and `theme_classic`. You've likely seen the default `ggplot2` theme *ad nauseam* at this point but let's show it one more time for comparative purposes.

```{r theme-default, warning = F, message = F}
core_plot
```

Now let's call the same plot but add `theme_bw` to it.

```{r theme-bw, warning = F, message = F}
core_plot +
  theme_bw()
```

Not bad! Finally, let's check out `theme_classic`.

```{r theme-classic, warning = F, message = F}
core_plot + 
  theme_classic()
```

Even better! If you'd like, you can use a pre-built `theme` *and* specify additional elements yourself. Let's add the `theme` content we came up with in the previous example to the `theme_classic` plot.

```{r theme-classic-manual, warning = F, message = F}
core_plot + 
  # Add the 'classic' theme
  theme_classic() +
  # As before, change the legend position and axis text size
  theme(legend.position = c(0.9, 0.15),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 14) )
```

Now that is looking publication-quality! One quick note though: **whichever theme specification is last "wins".** This means that if you have a `theme` call that sets font size to 14 and then add a second `theme` beneath it that sets font size to 30, the font size of your plot will be 30. This is part of why you can add pre-built themes and your own custom modifications together but is good to keep in mind if you're sure your custom theme is specified correctly but isn't showing up in your plot.
:::

### Challenge: `theme`

:::callout-important
## Your Turn!

Using the plot you created for the previous challenge (`bill_depth_mm` vs. `body_mass_g` and `color = sex`, etc.), (1) add the `theme_bw` pre-built theme, (2) increase the axis title size to 16, and (3) move the legend into the top left of the plot.
:::

### Combining `theme` and `+`

We discussed the strengths of `ggplot2`'s `+` operator earlier but it bears repeating here now that we've covered `theme`. An extremely helpful combination of these principles allows you to **save all of your theme editing into an object and then add that object to all the different plots you make for the project!** This would guarantee that all of your plots have the same format without you needing to re-write that formatting for every plot. We'll demonstrate briefly below.

### Example: `theme` & `+`

:::callout-note
## Example

Let's begin by defining our theme that we want all plots to use. For simplicity's sake, let's use the `theme` code outlined in the previous example.

```{r theme-plus-0, message = F, warning = F}
# Begin with `theme_classic` and then add on our custom modifications
my_theme <- theme_classic() +
  # As before, we're moving the legend and increasing axis font size
  theme(legend.position = c(0.7, 0.8),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 14) )
```

Now we can make a handful of plots and add the `my_theme` object to each of them!

```{r theme-plus-1, message = F, warning = F}
ggplot(penguins, aes(x = year, y = bill_depth_mm, color = sex)) +
  geom_jitter(width = 0.1) +
  labs(x = "Year", y = "Bill Depth (mm)",
       title = "Bill Depth x Year x Sex") +
  my_theme
```

```{r theme-plus-2, message = F, warning = F}
ggplot(penguins, aes(x = year, y = bill_length_mm, color = island)) +
  geom_jitter(width = 0.1) +
  labs(x = "Year", y = "Bill Length (mm)",
       title = "Bill Length x Year x Island") +
  my_theme
```

```{r theme-plus-3, message = F, warning = F}
ggplot(penguins, aes(x = year, y = flipper_length_mm, color = species)) +
  geom_jitter(width = 0.1) +
  labs(x = "Year", y = "Flipper Length (mm)",
       title = "Flipper Length x Year x Species") +
  my_theme
```

This can be a great way of making your figure creation process dramatically more efficient when it becomes time to shift from exploratory plots to publication-quality figures!
:::
